---
title: "practical 1"
---

## Excercise 1

1\.

```{r}
start_with_prefix <- function(string, prefix) {
  str_sub(string, 1, str_length(prefix)) == prefix
}



repeat_fill <- function(x, y) {
  rep(y, length.out = length(x))
}
#length(x) gets the size of x
#rep(y, length.out = ...) repeats y until it reaches that length
```

2.  Make a case for why `norm_r()`, `norm_d()` etc. would be better than `rnorm()`, `dnorm()`^[1](https://soda-lmu.github.io/StatProg2-2026-SoSe/02-functions/practical.html#fn1)^. Make a case for the opposite. How could you make the names even clearer?

-\> if you type "norm". you can easily get the variation of the data types.

## Excercise 2

```{r}
library(docstring)
library(palmerpenguins)
library(tidyverse)
library(ggplot2)
library(cli)

#' Displays the scatter plot of the selected columns on a selected species
#' @param species_name the name of the speciees  
#' @return GGplot

plot_species <- function(species_name, x_col = "bill_length_mm", y_col = "bill_depth_mm"){
  if (!any(penguins$species == species_name)) {
    
    stop("the selected species does not exist")
  }
  if (!is.numeric(dplyr::pull(penguins, {{ x_col }}))){
    
    cli_abort(c(
      "{.arg x_col} must be numeric"
    ))
  }
  if (!is.numeric(dplyr::pull(penguins, {{ y_col }}))){
    
    cli_abort(c(
      "{.arg y_col} must be numeric"
    ))
    
  }
  penguins %>%
    filter(species == species_name) %>%
    #use aes_string() to give the column name 
    ggplot(aes_string(x = x_col, y = y_col)) +
    geom_point() +
    labs(title = paste(x_col, "vs.", y_col, "for", species_name, "Penguins"),
         x = x_col,
         y = y_col)
}

plot_species("Adelie")

```

## **Exercise 4: Recursion and Iteration**

```{r}
#iterative version 
#' @param n the number of the numbers added.
fib_iter <- function(n) {
  if (n <= 0) stop("n must be a positive integer")
  # if there's only one number its 0
  if (n == 1) return(0)
  # if there's only two number it's 0 + 1 = 1
  if (n == 2) return(1)
  
  a <- 0
  b <- 1
  
  for (i in 3:n) {
    temp <- a + b
    a <- b
    b <- temp
  }
  
  return(b)
}

system.time({
  result1 <- fib_iter(30)
})

print(result1)
```

```{r}
#recursive version
fib_rec <- function(n) {
  if (n <= 0) stop("n must be a positive integer")
  if (n == 1) return(0)
  if (n == 2) return(1)
  
  fib_rec(n - 1) + fib_rec(n - 2)
}

system.time({
  result2 <- fib_rec(30)
})

print(result2)


```

## **Task 1: Spot the repetition**

Consider the following function that produces a small summary for a single species:

```{r}
summarise_species <- function(species_name) {
  adelie_mean_bill <- mean(penguins$bill_length_mm[penguins$species == "Adelie"], na.rm = TRUE)
  gentoo_mean_bill <- mean(penguins$bill_length_mm[penguins$species == "Gentoo"], na.rm = TRUE)
  chinstrap_mean_bill <- mean(penguins$bill_length_mm[penguins$species == "Chinstrap"], na.rm = TRUE)

  adelie_mean_mass <- mean(penguins$body_mass_g[penguins$species == "Adelie"], na.rm = TRUE)
  gentoo_mean_mass <- mean(penguins$body_mass_g[penguins$species == "Gentoo"], na.rm = TRUE)
  chinstrap_mean_mass <- mean(penguins$body_mass_g[penguins$species == "Chinstrap"], na.rm = TRUE)

  if (species_name == "Adelie") {
    cat("Adelie — mean bill:", adelie_mean_bill, "mean mass:", adelie_mean_mass, "\n")
  } else if (species_name == "Gentoo") {
    cat("Gentoo — mean bill:", gentoo_mean_bill, "mean mass:", gentoo_mean_mass, "\n")
  } else if (species_name == "Chinstrap") {
    cat("Chinstrap — mean bill:", chinstrap_mean_bill, "mean mass:", chinstrap_mean_mass, "\n")
  }
}

summarise_species("Adelie")
```

2\.

```{r}

summarise_species <- function(species_name){
  penguins %>%
    filter(species == species_name) %>%
    summarize(
      mean_bill = mean(bill_length_mm, na.rm = TRUE),
      mean_mass = mean(body_mass_g, na.rm = TRUE)
    ) %>%
    mutate(output = paste(species_name, " - mean bill: ", mean_bill, "mean bass: ", mean_mass, "\n")) %>%
    pull(output) %>%
    cat("\n")
}

summarise_species("Adelie")
```

### Task 2

```{r}
penguin_report <- function(species_name) {
  # load
  library(palmerpenguins)
  library(tidyverse)
  data <- penguins

  # clean
  data <- data[!is.na(data$bill_length_mm) & !is.na(data$body_mass_g), ]

  # filter
  data <- data[data$species == species_name, ]

  # summarise
  cat("n =", nrow(data), "\n")
  cat("mean bill length:", mean(data$bill_length_mm), "\n")
  cat("mean body mass:",   mean(data$body_mass_g),    "\n")

  # plot
  ggplot(data, aes(x = bill_length_mm, y = body_mass_g)) +
    geom_point() +
    labs(title = paste("Bill length vs. body mass —", species_name))
}

penguin_report("Gentoo")
```

```{r}

penguins_loading <- function(){
   # load
  library(palmerpenguins)
  library(tidyverse)
  data <- penguins
  
  return(data)
}

penguins_clean <- function(data){
  data <- data[!is.na(data$bill_length_mm) & !is.na(data$body_mass_g), ]
  
  return(data)
}

penguins_summarise_species <- function(data, species_name){
    # filter
  data <- data[data$species == species_name, ]

  # summarise
  cat("n =", nrow(data), "\n")
  cat("mean bill length:", mean(data$bill_length_mm), "\n")
  cat("mean body mass:",   mean(data$body_mass_g),    "\n")
  
  return(data)
}

penguins_plot <- function(data, species_name){
    ggplot(data, aes(x = bill_length_mm, y = body_mass_g)) +
      geom_point() +
      labs(title = paste("Bill length vs. body mass —", species_name))
}

penguin_report <- function(species_name){
  df <- penguins_loading()
  df <- penguins_clean(df)
  df <- penguins_summarise_species(df, species_name)
  penguins_plot(df, species_name)
}
  

penguin_report("Gentoo")
```

```{r}

library(tidyverse)
library(ggplot2)

#' clean the columns used for the plot later
starwars_clean <- function(data, column1, column2){
  data <- data %>%
    filter(!is.na(data[[column1]]), !is.na(data[[column2]]))
  
  return(data)
}

starwars_means <- function(data, column1, column2){
  cat(column1, "mean: ", mean(data[[column1]]), "\n")
  cat(column2, "mean: ", mean(data[[column2]]), "\n")
  
  return(data)
}

starwars_plot <- function(data, column1, column2){
  ggplot(data, aes_string(x = column1, y = column2)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    xlab(column1) + 
    ylab(column2) + 
    labs(
    title = paste(column1, " vs ", column2)
  ) 
    
    
}

df <- starwars
starwars_report <- function(df, column1, column2){
  df <- starwars_clean(df, column1, column2)
  df <- starwars_means(df, column1, column2)
  starwars_plot(df, column1, column2)
}

starwars_report(df, "height", "mass")
```
