---
title: "Project Outline"
format: html
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, message=FALSE,warning=FALSE}

#load packages
library(tidyverse)
library(ggplot2)
library(plotly)
library(readxl)
library(broom)
library(rpart)
library(rpart.plot)

#load data
happiness <- read_excel("world_hapiness_datasets/WHR25_Data_Figure_2.1v3.xlsx")

#clean data
happiness_clean <- happiness %>%
  rename(
    year = Year,
    rank = Rank,
    country = `Country name`,
    happiness = `Life evaluation (3-year average)`,
    gdp = `Explained by: Log GDP per capita`,
    social_support = `Explained by: Social support`,
    health = `Explained by: Healthy life expectancy`,
    freedom = `Explained by: Freedom to make life choices`,
    generosity = `Explained by: Generosity`,
    corruption = `Explained by: Perceptions of corruption`,
    dystopia_residual = `Dystopia + residual`
  ) %>%
  mutate(across(c(happiness, gdp, social_support, health, freedom,
                  generosity, corruption, dystopia_residual), as.numeric))

#filter complete cases
happiness_complete <- happiness_clean %>%
  filter(
    !is.na(happiness),
    !is.na(gdp),
    !is.na(social_support),
    !is.na(health),
    !is.na(freedom),
    !is.na(generosity),
    !is.na(corruption)
  )
```

Every March, the World Happiness Report turns well-being into a global headline: Finland is once again ranked as the happiest country in the world. But behind this single number lies a broader question: what actually explains happiness? Is it mainly economic prosperity, or do social support, health, freedom, and trust matter just as much?

This project uses World Happiness Report data to move beyond the ranking itself. Instead of only asking which countries are happiest, we investigate which factors are most strongly associated with happiness, where countries form similar well-being profiles, and which parts of happiness remain unexplained by the available variables.

# Initial data analysis

Before modelling anything, we look at what the raw data actually contains.

**What a row is.** Each row is one **country observed in one report year**, so the pair `(country, year)` uniquely identifies an observation. The raw file `WHR25_Data_Figure_2.1v3.xlsx` holds `r nrow(happiness)` rows spanning `r min(happiness$Year)`–`r max(happiness$Year)` and `r n_distinct(happiness$"Country name")` countries.

**Missingness.** The six explanatory factors are not reported for every country-year. Generosity and perceptions of corruption in particular are missing for many of the earlier years:

```{r ida-missing}
happiness_clean %>%
  summarise(across(
    c(happiness, gdp, social_support, health, freedom, generosity, corruption),
    ~ sum(is.na(.))
  )) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing values") %>%
  arrange(desc(`Missing values`)) %>%
  knitr::kable(caption = "Missing values per variable (raw data)")
```

**Effect of the complete-case filter.** For the analysis we keep only rows with no missing values across the happiness score and all six factors. This is a consequential step, not a formality:

```{r ida-dropped}
tibble(
  Stage = c("Raw rows", "Complete cases kept", "Rows dropped"),
  n = c(nrow(happiness_clean), nrow(happiness_complete),
        nrow(happiness_clean) - nrow(happiness_complete))
) %>%
  knitr::kable(caption = "Effect of the complete-case filter")
```

Filtering to complete cases removes `r nrow(happiness_clean) - nrow(happiness_complete)` of `r nrow(happiness_clean)` rows (about `r round(100 * (nrow(happiness_clean) - nrow(happiness_complete)) / nrow(happiness_clean))`%). Because generosity and corruption are unavailable in the earliest years, this step also narrows the effective sample to **`r min(happiness_complete$year)`–`r max(happiness_complete$year)`**. Every plot and model below is therefore built on the remaining **`r nrow(happiness_complete)` complete country-year observations** covering `r n_distinct(happiness_complete$country)` countries.

**Distribution of the target.** Happiness scores are roughly bell-shaped and centred just above the middle of the 0–10 Cantril ladder, with a slight left skew — very few countries score below 3:

```{r ida-dist}
ggplot(happiness_complete, aes(x = happiness)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of happiness scores",
    subtitle = "Complete country-year observations used in the analysis",
    x = "Life evaluation / happiness score (0–10)",
    y = "Count"
  ) +
  theme_minimal()
```

# Research Questions

## Q1. What explains happiness?

**To what extent can happiness be explained by economic and social factors such as GDP, social support, and freedom?**

This question explores whether richer countries are automatically happier, or whether other dimensions such as social support and freedom help explain why some countries achieve higher well-being than others.

------------------------------------------------------------------------

### Plot 1: GDP and happiness {#plot1}

This plot compares GDP contribution with happiness. A linear regression line shows the general direction of the relationship, while the LOESS curve allows for a non-linear pattern. This matters because the relationship between income and happiness is not constant: as the LOESS curve below confirms, the gains flatten at higher income levels.

```{r warning=FALSE, message=FALSE}
create_hover <- function(country, year, x_value, happiness, label) {
  paste(
    "Country:", country,
    "<br>Year:", year,
    paste0("<br>", label, ":"), round(x_value, 2),
    "<br>Happiness score:", round(happiness, 2)
  )
}
```

```{r warning=FALSE, message=FALSE}
#This is plot.

p_gdp <- ggplot(
  happiness_complete,
  aes(
    x = gdp,
    y = happiness,
    text = create_hover(
      country,
      year,
      gdp,
      happiness,
      "GDP contribution"
      )
    )
  )+
  geom_point(alpha = 0.45, color = "gray40") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "darkblue",
    linewidth = 1,
    aes(text = NULL)
  ) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    color = "red",
    linewidth = 1,
    aes(text = NULL)
  ) +
  labs(
    title = "GDP and Happiness: Linear Trend vs Nonlinear Pattern",
    subtitle = "The LOESS curve helps reveal possible diminishing returns of GDP",
    x = "Explained by: Log GDP per capita",
    y = "Life Evaluation / Happiness Score"
  ) +
  theme_minimal()

ggplotly(p_gdp, tooltip = "text")
```

**Interpretation:**\
The scatter plot shows a clear positive relationship between GDP per capita and happiness scores. Countries with higher GDP contributions generally tend to report higher levels of happiness.

The blue linear regression line indicates that happiness increases steadily as GDP increases: the fitted line rises by about **1.67 happiness points for each one-unit increase in the log-GDP contribution** (correlation *r* ≈ 0.69). This suggests that economic prosperity is an important factor associated with higher well-being.

The red LOESS curve reveals that the relationship is not perfectly linear. At very low GDP levels, increases in GDP are associated with relatively small changes in happiness. In the middle range, the relationship becomes steeper, indicating that increases in GDP are accompanied by larger gains in happiness. At higher GDP levels, the curve begins to flatten slightly, suggesting that additional income may produce smaller improvements in happiness than at lower income levels.

The noticeable spread of observations around both trend lines indicates that GDP alone does not fully explain happiness differences across countries. Countries with similar GDP levels can display quite different happiness scores, implying that other factors such as social support, health, freedom, and governance also contribute to well-being.

### Plot 2: Social support and happiness {#plot2}

This plot focuses on the relationship between social support and happiness. It directly supports the central story of the project: well-being is not only about wealth, but also about relationships and social connection.

```{r warning=FALSE}
# This is a plot.

p_social <- ggplot(
  happiness_complete,
  aes(
    x = social_support,
    y = happiness,
    text = create_hover(
      country,
      year,
      gdp,
      happiness,
      "Social support"
      )
    )
  )+
  geom_point(alpha = 0.45, color = "gray40") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "darkgreen",
    linewidth = 1,
    aes(text = NULL)
    ) +
  labs(
    title = "Social Support and Happiness",
    subtitle = "Countries with stronger social support tend to report higher happiness",
    x = "Explained by: Social support",
    y = "Life Evaluation / Happiness Score"
    ) +
  theme_minimal()
ggplotly(p_social, tooltip = "text")
```

**Interpretation:**\
The plot shows a clear **positive relationship** between social support and happiness. Countries with higher levels of social support generally report higher happiness scores, as indicated by the upward-sloping regression line.

The relatively strong upward trend — a slope of about **2.18 happiness points per unit of social-support contribution** (correlation *r* ≈ 0.69, essentially the same strength as GDP) — suggests that social support is an important predictor of well-being. Countries where people can rely on friends, family, or their community in times of need tend to have higher average happiness levels.

Although the relationship is clearly positive, there is still noticeable variation around the regression line. Countries with similar levels of social support do not always achieve the same happiness scores, indicating that other factors such as income, health, freedom, or governance also contribute to differences in well-being.

Compared with the GDP plot, this figure demonstrates that economic factors are not the only determinants of happiness. Strong social connections appear to be closely associated with higher happiness levels across countries.

### Plot 3: Which factors are most strongly associated with happiness? {#plot3}

Instead of looking at only one variable at a time, this plot compares the correlation between happiness and all explanatory factors. It summarizes which variables are most closely related to happiness.

```{r warning=FALSE}
#This is a calculation + plot.It is not a predictive model, but it calculates correlations and then plots them.

cor_data <- happiness_complete %>%
  select(happiness, gdp, social_support, health, freedom, generosity, corruption)

correlations <- cor_data %>%
  summarise(
    GDP = cor(happiness, gdp),
    `Social Support` = cor(happiness, social_support),
    Health = cor(happiness, health),
    Freedom = cor(happiness, freedom),
    Generosity = cor(happiness, generosity),
    Corruption = cor(happiness, corruption)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "factor",
    values_to = "correlation"
  ) %>%
  arrange(desc(correlation))

ggplot(correlations, aes(x = reorder(factor, correlation), y = correlation)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Correlation with Happiness",
    subtitle = "Comparison of the reported explanatory factors",
    x = "Factor",
    y = "Correlation with happiness"
  ) +
  theme_minimal()
```

**Interpretation:**\
The bar chart shows the correlation between happiness and each explanatory factor. Larger correlation values indicate a stronger association with happiness.

The strongest correlations are observed for GDP (*r* ≈ 0.69) and Social Support (*r* ≈ 0.69), both close to 0.7. This suggests that countries with higher income levels and stronger social support networks tend to report higher happiness scores.

Health also exhibits a strong positive correlation with happiness (*r* ≈ 0.66), indicating that longer and healthier lives are associated with greater well-being.

Freedom has a moderate positive correlation (*r* ≈ 0.54), suggesting that the ability to make life choices is linked to higher happiness, although its relationship is weaker than those of GDP, social support, and health.

Corruption shows a weaker but still meaningful correlation with happiness (*r* ≈ 0.43). This implies that perceptions of lower corruption are associated with higher well-being, although the relationship is less pronounced.

In contrast, Generosity is essentially uncorrelated with happiness (*r* ≈ 0.04). Within this dataset, generosity appears to be much less strongly associated with happiness than the other explanatory factors.

### Plot 4: Country profiles using clustering {#plot4}

This plot uses k-means clustering to group countries with similar happiness profiles. It serves Q1 directly: instead of testing one factor at a time, it asks whether countries fall into distinct well-being *profiles* across all six factors at once. Unlike regression, clustering does not predict happiness. Instead, it discovers patterns in the data and groups countries based on similarities across multiple factors. We cluster the `r sum(happiness_complete$year == 2024)` countries with complete 2024 data.

```{r warning=FALSE}
#This is a model + plot
#kmeans() = clustering model
#ggplot() = visualization of the cluster results

cluster_data <- happiness_complete %>%
  filter(year == 2024) %>%
  select(country, happiness, gdp, social_support, health, freedom, generosity, corruption)

cluster_scaled <- cluster_data %>%
  select(gdp, social_support, health, freedom, generosity, corruption) %>%
  scale()

set.seed(123)
clusters <- kmeans(cluster_scaled, centers = 3)

cluster_data <- cluster_data %>%
  mutate(
    cluster = as.factor(clusters$cluster),
    hover_text = paste(
      "Country:", country,
      "<br>Cluster:", cluster,
      "<br>Happiness score:", round(happiness, 2),
      "<br>GDP contribution:", round(gdp, 2),
      "<br>Social support:", round(social_support, 2),
      "<br>Health:", round(health, 2),
      "<br>Freedom:", round(freedom, 2)
    )
  )

p_cluster <- ggplot(
  cluster_data,
  aes(
    x = gdp,
    y = social_support,
    color = cluster,
    text = hover_text
  )
) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Country Clusters Based on Happiness Factors",
    subtitle = "k-means clustering using GDP, social support, health, freedom, generosity, and corruption",
    x = "Explained by: Log GDP per capita",
    y = "Explained by: Social support",
    color = "Cluster"
  ) +
  theme_minimal()

ggplotly(p_cluster, tooltip = "text")
```

**Interpretation:**\
The k-means clustering algorithm identifies three distinct groups of countries based on multiple happiness-related factors. Although the clusters were created using several variables, they are visualized here using GDP per capita (x-axis) and social support (y-axis).

The plot shows a clear pattern: countries with higher GDP levels generally also tend to have higher levels of social support. The clusters largely follow this gradient.

Cluster 1 (red) contains the largest group — about 46 countries — with relatively low GDP and lower social support and the lowest mean happiness (≈ 4.4). These countries tend to exhibit less favorable conditions associated with happiness.

Cluster 2 (green), roughly 67 countries, represents intermediate levels of GDP and social support with a mean happiness of about 5.9. This group can be viewed as a middle category between lower- and higher-performing countries.

Cluster 3 (blue), the smallest group at about 31 countries, contains those with the highest GDP and strongest social support and the highest mean happiness (≈ 6.8). These countries generally have characteristics associated with higher happiness levels.

The visible separation between clusters suggests that countries are not randomly distributed but form distinct groups with similar socioeconomic and well-being profiles.

### Q1 Summary

The Q1 analysis shows that happiness is associated with several dimensions. GDP matters, but the relationship is not the full story. Social support, freedom, health, and institutional trust also contribute to differences in happiness. The clustering model further suggests that countries form different well-being profiles rather than following one single pattern.

------------------------------------------------------------------------

## Q2. What remains unexplained?

**What part of happiness cannot be explained by observable variables?**

This question uses modeling to estimate happiness from the reported factors and then examines the residuals. Residuals are the differences between actual and predicted happiness. They help identify countries that are happier or less happy than expected based on the model.

### Plot 1: Predicted vs actual happiness {#plot5}

A multiple linear regression model predicts happiness from the six explanatory factors. The plot compares predicted happiness with actual happiness; the closer the points sit to the diagonal, the better the fit. Here they cluster tightly around it — the model reproduces about **71% of the variation in happiness (R² ≈ 0.71)** — though, as noted below, part of that tightness is built into how the factors are defined.

```{r q2-setup, include=FALSE}
# Prepare model data once for all Q2 plots

model_data <- happiness_complete %>%
  select(country, year, happiness, gdp, social_support, health, freedom, generosity, corruption)

# Fit multiple linear regression model
lm_model <- lm(
  happiness ~ gdp + social_support + health + freedom + generosity + corruption,
  data = model_data
)

# Create predictions and residuals
model_results <- model_data %>%
  mutate(
    predicted = predict(lm_model),
    residual = happiness - predicted
  )
```

```{r warning=FALSE, message=FALSE}
#This is a model + plot.
#lm() = linear regression model
#ggplot() = model performance visualization

# create hover text
model_results <- model_results %>%
  mutate(
    hover_text = paste(
      "Country:", country,
      "<br>Year:", year,
      "<br>Actual happiness:", round(happiness, 2),
      "<br>Predicted happiness:", round(predicted, 2),
      "<br>Residual:", round(residual, 2)
    )
  )

# build plot
p_pred <- ggplot(
  model_results,
  aes(
    x = predicted,
    y = happiness,
    text = hover_text
  )
) +
  geom_point(color = "gray40", alpha = 0.6) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = "red",
    linewidth = 1,
    aes(text = NULL)
  ) +
  labs(
    title = "Predicted vs Actual Happiness",
    subtitle = "Multiple linear regression using six reported factors",
    x = "Predicted happiness",
    y = "Actual happiness"
  ) +
  theme_minimal()

# interactive plot
ggplotly(p_pred, tooltip = "text")
```

**Interpretation:**\
The plot compares the predicted happiness scores from your multiple linear regression model (x-axis) with the actual happiness scores from the dataset (y-axis).

The red diagonal line represents perfect predictions. If the model predicted happiness perfectly, all points would lie exactly on this line.

In this plot, most observations cluster around the diagonal line, indicating that the model captures a substantial portion of the variation in happiness scores — the six factors together explain about **71% of the variance in happiness (R² ≈ 0.71)**. This suggests that the explanatory variables included in the model are useful predictors of happiness.

It is worth being cautious here, though. As noted on the [dataset page](data.qmd), the six "Explained by" columns are *contributions* that, together with the Dystopia + residual term, sum to the happiness score by construction. Regressing the score on these contributions is therefore partly circular, which is one reason the fit looks as tight as it does. The residuals below are more informative than the R² itself.

However, there is still noticeable scatter around the line. Some countries have actual happiness scores that are higher or lower than the model predicts, meaning that prediction errors remain. This indicates that the model does not fully explain all differences in happiness and that other factors not included in the regression may also influence happiness levels.

The spread of points appears relatively consistent across the range of predictions, and there is no obvious systematic pattern of over- or underestimation. Overall, the model demonstrates a reasonably good fit, although prediction accuracy is not perfect.

### Plot 2: Countries that deviate from the model {#plot6}

This plot shows the strongest positive and negative residuals. Positive residuals mean that a country is happier than predicted by the model. Negative residuals mean that a country is less happy than predicted.

Residual=Actual happiness−Predicted happiness

```{r warning=FALSE}
#This is a model-result plot.It does not fit a new model, but it visualizes the residuals from the previous regression model.

residual_summary <- model_results %>%
  group_by(country) %>%
  summarise(mean_residual = mean(residual, na.rm = TRUE)) %>%
  ungroup()

top_residuals <- residual_summary %>%
  slice_max(mean_residual, n = 8)

bottom_residuals <- residual_summary %>%
  slice_min(mean_residual, n = 8)

residual_plot_data <- bind_rows(top_residuals, bottom_residuals) %>%
  mutate(type = if_else(mean_residual > 0, "Happier than predicted", "Less happy than predicted"))

ggplot(residual_plot_data, aes(x = reorder(country, mean_residual), y = mean_residual, fill = type)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Countries That Deviate Most from the Regression Model",
    subtitle = "Average residuals across available years",
    x = "Country",
    y = "Average residual",
    fill = "Residual type"
  ) +
  theme_minimal()
```

**Interpretation:**\
The chart shows the countries with the largest average residuals across all years.

Countries on the right side (positive residuals), such as Guatemala, Côte d'Ivoire, Costa Rica, Israel, and Finland, are consistently happier than the model predicts. This suggests that factors beyond those included in the regression model may contribute positively to happiness in these countries.

In contrast, countries on the left side (negative residuals), such as Rwanda, Sri Lanka, Botswana, Lebanon, and Hong Kong SAR of China, are less happy than predicted by the model. Although these countries have characteristics that would lead the model to expect higher happiness scores, their observed happiness levels are lower, indicating that additional unobserved factors may negatively affect well-being.

The fact that residuals range from approximately −1.5 to +0.9 happiness points shows that while the model performs reasonably well overall, it does not fully capture all determinants of happiness. Some countries systematically deviate from the model's predictions.

### Plot 3: Decision tree model {#plot7}

The decision tree provides a different modeling perspective. Unlike linear regression, it splits the data into groups based on factor thresholds. This makes it useful for identifying which variables separate countries with lower and higher happiness scores.

```{r warning=FALSE}
#This is a model / model visualization.
#rpart() = decision tree model
#rpart.plot() = visualization of the model structure

tree_model <- rpart(
  happiness ~ gdp + social_support + health + freedom + generosity + corruption,
  data = model_data,
  method = "anova",
  control = rpart.control(cp = 0.01)
)

rpart.plot(
  tree_model,
  main = "Decision Tree Predicting Happiness",
  type = 4,
  extra = 101,
  fallen.leaves = TRUE
)
```

**Interpretation:**\
The decision tree predicts happiness by repeatedly splitting countries into groups based on threshold values of explanatory variables.

The first and most important split is health (`health < 0.5`). This means that health is the strongest factor in the tree for distinguishing between countries with lower and higher happiness scores. Countries with poorer health tend to fall into branches with lower predicted happiness values.

Among countries with lower health levels, social support, freedom, and GDP provide additional splits that further separate countries into different happiness groups. For example, countries with low health and low social support are predicted to have some of the lowest happiness scores in the dataset.

For countries with better health, the next important factor is corruption, followed by social support, freedom, and GDP. The highest predicted happiness scores occur in groups that combine good health, low perceived corruption, and relatively high freedom.

Overall, the tree suggests that health, social support, freedom, GDP, and corruption are important determinants of happiness, with health appearing as the most influential variable because it is selected for the first split of the tree.

## Final Conclusion

The analysis shows that happiness is not determined by wealth alone. GDP is clearly associated with happiness, but social support, freedom, health, and institutional trust also play important roles. The clustering model shows that countries follow different happiness profiles, while the regression and decision tree models show that some patterns can be modeled but not fully explained.

Overall, the project supports the idea that economic prosperity matters, but it cannot fully capture human well-being. The residuals show that there is still an unexplained part of happiness—possibly shaped by culture, inequality, mental health, social stability, or other factors not directly measured in the dataset.

**Main takeaway:**\
Wealth matters, but happiness is broader than money.