```{r}
library(tidyverse)
library(readxl)
Our_Data_ <- read_excel("C:/Users/Z005805K/Downloads/Eya/Stat/Stat_Project_Attempt/Stat/Quarto/world_hapiness_datasets/WHR25_Data_Figure_2.1v3.xlsx")
View(Our_Data_)
```

---
title: "Project Proposal"
format: html
---

# 1. Project Outline

Every March, headlines announce that Finland is “the happiest country in the world” — again. But what actually sits behind that single number, how much does it move from year to year, and how much of a nation's happiness can be pinned to money, health and freedom? This project uses five years of global survey data to look past the headline ranking and ask what really travels with well-being.

## Member

\[name\] \[email\]

\[name\] \[email\]

Shohei Kobayashi (kobayashishohei1023\@gmail.com)

## GitHub Repogitry

<https://github.com/Youta-droid/Stat_Project>

# 2. Source & Licence

**Dataset:** World Happiness Report

**Hosted on:** Kaggle — [kaggle.com/datasets/unsdsn/world-happiness](https://www.kaggle.com/datasets/unsdsn/world-happiness)

**Published by:** UN Sustainable Development Solutions Network (UNSDSN)

**Underlying source:** Gallup World Poll — the Cantril ladder life-evaluation question

**Coverage:** Five annual files, 2015–2019 (one CSV per report year)

**Licence:** CC0 1.0 — Public Domain (no rights reserved)

Because the data is released under CC0, there is no legal obligation to attribute it. As good academic practice we will still credit it, e.g. “World Happiness Report, UNSDSN, via Kaggle (CC0). Underlying data: Gallup World Poll.”

# 3. What's in It

Each row represents one country in one report year. Combined across the five files there are roughly 780 rows and 150–160 countries per year. Crucially, the schema is not identical across years — column counts and names drift — so a small harmonisation step is needed before any cross-year analysis.

## Per-year structure

+----------+------------------------------+----------+-------------------------------------------------------------------------------+
| **Year** | **Rows**                     | **Cols** | **What is different that year**                                               |
+----------+------------------------------+----------+-------------------------------------------------------------------------------+
| **2015** | 158                          | 12       | Includes Region and Standard Error; has Dystopia Residual                     |
+----------+------------------------------+----------+-------------------------------------------------------------------------------+
| **2016** | 157                          | 13       | Includes Region; Standard Error replaced by Lower/Upper Confidence Interval   |
+----------+------------------------------+----------+-------------------------------------------------------------------------------+
| **2017** | 155                          | 12       | Region dropped; Whisker high/low instead of CI; Dystopia Residual kept        |
+----------+------------------------------+----------+-------------------------------------------------------------------------------+
| **2018** | 156                          | 9        | Slim schema: columns renamed; Region, CI and Dystopia Residual all removed    |
+----------+------------------------------+----------+-------------------------------------------------------------------------------+
| **2019** | 156                          | 9        | Same slim schema as 2018                                                      |
+----------+------------------------------+----------+-------------------------------------------------------------------------------+

Row counts vary because the set of surveyed countries changes slightly each year; column counts vary because the report's format was simplified over time.

## Key variables

+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Concept**             | **Column name(s)  (2015–17  /  2018–19)**                     | **What it measures**                                         |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Country**             | Country  /  Country or region                                 | The nation — the unit of analysis                            |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Region**              | Region  (2015–2016 only)                                      | World region grouping (e.g. Western Europe)                  |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Rank**                | Happiness Rank  /  Overall rank                               | Position in that year's global ranking                       |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Happiness score ★**   | Happiness Score  /  Score                                     | Cantril-ladder life evaluation, 0–10 — the TARGET variable   |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **GDP per capita**      | Economy (GDP per Capita)  /  GDP per capita                   | Contribution of income to the score                          |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Social support**      | Family  /  Social support                                     | Contribution of having someone to count on                   |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Life expectancy**     | Health (Life Expectancy)  /  Healthy life expectancy          | Contribution of healthy longevity                            |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Freedom**             | Freedom  /  Freedom to make life choices                      | Contribution of freedom over life choices                    |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Generosity**          | Generosity                                                    | Contribution of charitable behaviour                         |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Corruption**          | Trust (Government Corruption)  /  Perceptions of corruption   | Contribution of perceived honesty / low corruption           |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+
| **Dystopia residual**   | Dystopia Residual  (through 2017)                             | Unexplained remainder above the Dystopia baseline            |
+-------------------------+---------------------------------------------------------------+--------------------------------------------------------------+

★ = target variable. Where two names are shown, the report renamed the column in the 2018–2019 files.

## A subtlety that shapes the whole analysis

**The six “factor” columns are not raw measurements.**

**What they actually are:** Each factor column (GDP, social support, life expectancy, freedom, generosity, corruption) is the estimated contribution of that factor to the country's score — not the raw value (e.g. it is not GDP in dollars).

**They are additive:** The six contributions plus the Dystopia Residual sum (up to rounding) to the Happiness Score. The bars in the report literally stack to the total.

**Why it matters:** This means a model that “predicts” the score from these six columns will score near-perfectly almost by construction — a fact we deliberately exploit (and flag) in Question 3.

# 4. Research Questions (Updated)

## Q1. How strongly are GDP and social support associated with happiness, and does GDP show diminishing returns?

**Planned output:**
- Scatter plot: GDP per capita vs. happiness (with LOESS curve)
#GDP vs. happiness
```{r}
ggplot(Our_Data_, aes(
  x = `Explained by: Log GDP per capita`,
  y = `Life evaluation (3-year average)`
)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", color = "blue") +
  labs(
    title = "GDP vs Happiness with LOESS Curve",
    x = "GDP per Capita",
    y = "Happiness Score"
  )
```

```{r}
ggplot(Our_Data_, aes(x = `Explained by: Log GDP per capita`, y = `Life evaluation (3-year average)`)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", color = "blue") +
  labs(
    title = "GDP vs Happiness with LOESS Curve",
    x = "GDP per Capita",
    y = "Happiness Score"
  )
```
  
- Scatter plot: social support vs. happiness
#social support vs.happiness
```{r}
ggplot(Our_Data_, aes(x = `Explained by: Social support`, y = `Life evaluation (3-year average)`)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", color = "red") +
  labs(
    title = "Social Support vs Happiness",
    x = "Social Support",
    y = "Happiness Score"
  )
```

- Correlation comparison (standardised effect sizes)


**Claim type:** Exploratory

**Goal:**
To assess whether GDP and social support are both strongly associated with happiness, and whether the effect of GDP flattens at higher income levels.

---

## Q2. How well do the six reported factors explain happiness scores, and which countries deviate from the model?

**Planned output:**
- Multiple linear regression model (happiness ~ six factors)
- Model performance: R² and RMSE
- Ranking of countries by residuals (over- and under-performers)

**Claim type:** Predictive

**Goal:**
To quantify how much of happiness can be explained by the given factors and identify countries that are significantly above or below model expectations.

# 5. Context & Target Audience

**Context:** Journalistic — a data-driven story for an educated general audience curious about global well-being, not a peer-reviewed paper.

**Audience:** General readers, students and the policy-curious public; no statistical background assumed.

## Candidate story angle

“Rich, free — and still restless: what five years of global happiness data reveal about the part of well-being money can't buy.” Headline ranking as the hook (Q1), the wealth-vs-relationships tension as the body (Q2), and the unexplained residual as the twist (Q3).

## Presentation formats (mapped to each question)

• **Choropleth world map —** happiness scores by country, shaded by score range — sets the scene for Q1.

• **Slope / line chart —** top-10 countries' scores across 2015–2019, plus a small “biggest movers” table — the evidence for Q1.

• **Annotated scatter —** GDP per capita vs. happiness with a smoothed fit, plus a correlation heatmap — carries Q2.

• **Coefficient & residual chart —** which factors weigh most, and a ranked list of over- / under-performers by residual — delivers Q3.

• **Interactive dashboard —** lets readers filter by year or region and explore on their own — ties the three threads together.

*These formats prioritise accessibility and visual storytelling over statistical formalism, which fits a journalistic audience that wants the story behind the numbers.*

# Proposed Next Step

Load the five CSV files in R and harmonise them into one tidy data frame: add a Year column, standardise the differing column names, and decide how to handle the columns that exist only in some years (Region, the confidence-interval fields, and the Dystopia Residual). After that, begin exploratory analysis for Q1.
